Spatial Investigation and Exploration of Factors influencing HIV-related Mortality

### 2023-04-30 code/analysis_code/p4_mapping.Rmd
Auto stash before revert of "fixed pca and models"

df_scaled <- readRDS(here("data/processed_data/df_fit_alt.rds"))
df_raw <- readRDS(here("data/processed_data/race_alt.rds"))
```

I do a bit more cleaning to the data to tidy things up and create some new variables
```{r}
df_scaled <- df_scaled %>%
  mutate(region = tolower(State)) %>% 
  replace(is.na("Ratio Observed/ Expected"), 0) %>% 
  mutate(Exp_DR = `Expected Death Cases`/Population *1e5,
               OE_mortality_ratio = `Death Rate (estimated)`/Exp_DR,
         dykes_chi = ((`Death Cases`- `Expected Death Cases`)/ sqrt(`Expected Death Cases`)),

         ## GLM
         predicted_observed_mult = mult_fit/`MRR`,
         predicted_observed_mult = ifelse(!is.finite(predicted_observed_mult), 0, predicted_observed_mult),
         observed_predicted_mult = `Death Cases`/mult_fit,
         observed_predicted_mult = ifelse(!is.finite(observed_predicted_mult), 0, observed_predicted_mult),
         ## CAR
         predicted_observed_car = car_fit/`MRR`,
         predicted_observed_car = ifelse(!is.finite(predicted_observed_car), 0, predicted_observed_car),
         observed_predicted_car = `Death Cases`/car_fit,
         observed_predicted_car = ifelse(!is.finite(observed_predicted_car), 0, observed_predicted_car),
         ## POISSON
         predicted_observed_pois = pois_fit/`MRR`,
         predicted_observed_pois = ifelse(!is.finite(predicted_observed_pois), 0, predicted_observed_pois),
         observed_predicted_pois = `Death Cases`/pois_fit,
         observed_predicted_pois = ifelse(!is.finite(observed_predicted_pois), 0, observed_predicted_pois),
         ##N NEGATIVE BINOMIAL
         predicted_observed_nb = nb_fit/`MRR`,
         predicted_observed_nb = ifelse(!is.finite(predicted_observed_nb), 0, predicted_observed_nb),
         observed_predicted_nb = `Death Cases`/nb_fit,
         observed_predicted_nb = ifelse(!is.finite(observed_predicted_nb), 0, observed_predicted_nb))

df_raw <- df_raw %>%
  mutate(region = tolower(State))
```

Next, I will merge my dataframe contain the variables of interest with the spatial object. 
```{r}
df_sf_scale <- merge(states,df_scaled, by = "region")
# This next step is import to ensure that no "holes" appear in the map.
df_sf_scale <- df_sf_scale[order(df_sf_scale$order), ]

df_sf_raw <- merge(states,df_raw, by = "region")
# This next step is import to ensure that no "holes" appear in the map.
df_sf_raw <- df_sf_raw[order(df_sf_raw$order), ]
```

Now I'm ready to start mapping! I'll begin by looking at the map of the raw HIV death rate across to contiguous US by selected race/ ethnicity.
```{r}
ggplot(df_sf_scale, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Death Rate (estimated)`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race)  + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "HIV Mortality Rate (raw) by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Death rate per 100,000") +
  scale_fill_distiller(palette = "RdYlGn") +
  theme(plot.title= element_text(hjust = 0.5))

@@

```{r}
ggplot(df_sf_scale, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`MRR`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio in the United States, 2020", x = "", y = "", fill = "Ratio") +
  scale_fill_distiller(palette = "RdYlGn") +
  theme(plot.title= element_text(hjust = 0.5))

@@

```{r}
ggplot(df_sf_raw, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Gini Coefficient`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Gini Coefficient, 2020", x = "", y = "", fill = "Gini coefficient (higher values represent greater income inequality)")  +
  theme(plot.title= element_text(hjust = 0.5))
ggsave(here("results/gini_map.png"), units="in", width=10, height=7, dpi=300)
```

```{r}
ggplot(df_sf_raw, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Percent Uninsured Under 65 Years Old`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Percent Uninsured Under 65 Years Old, 2020", x = "", y = "", fill = "%") +
  theme(plot.title= element_text(hjust = 0.5)) 
ggsave(here("results/uninsured_map.png"), units="in", width=10, height=7, dpi=300)
```

```{r}
ggplot(df_sf_raw, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Percent Without an Internet subscription`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Percent Without an Internet Subscription, 2020", x = "", y = "", fill = "%") +
  theme(plot.title= element_text(hjust = 0.5)) 
ggsave(here("results/no_internet_map.png"), units="in", width=10, height=7, dpi=300)
```

@@

Next, let's map a probability map based on where we'd expect deaths if risk of HIV-related death was consistent across the US.
```{r}
ggplot(df_sf_scale, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = Exp_DR)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race) +
  scale_fill_viridis_c() + theme_void() + 
  theme(strip.text = element_text(size = 14),
    strip.background = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.text = element_blank(),
        legend.position = "bottom") +
  labs(title = "NULL - Expected HIV Mortality Rate by Race/ Ethnicity
       in the United States, 2020", x = "", y = "", fill = "Death Rate (per 100,000)") +
  theme(plot.title= element_text(hjust = 0.5))
```

Next, let's using Dykes and Unwin's method of smooth map using a chi square value. 
```{r}
ggplot(df_sf_scale, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = dykes_chi)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race)  + theme_void() + 
  theme(strip.text = element_text(size = 14),
        strip.background = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.text = element_blank(),
        legend.position = "bottom") +
  labs(title = "Dykes and Unwin's Chi Square by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Chi Square") +
  scale_fill_distiller(palette = "RdYlGn") +
  theme(plot.title= element_text(hjust = 0.5))

@@

```{r}
##Smoothed map using fitted values from best model (CAR)
ggplot(df_sf_scale, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = car_fit)) +
  scale_fill_gradientn(colors = rev(brewer.pal(9, "RdYlGn"))) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)  + 
  theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Smoothed Map of Black-to-White HIV-Related Mortality Rate Ratio,
United States, 2020 (CAR)", x = "Ratio", y = "", fill = "") +
  theme(plot.title= element_text(hjust = 0.5))

ggsave(here("results/car_fit_map.png"), units="in", width=10, height=4, dpi=300)
```

```{r}
##Smoothed map using fitted values from MULT (only marginally worse than CAR)
ggplot(df_sf_scale, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = mult_fit)) +
  scale_fill_gradientn(colors = rev(brewer.pal(9, "RdYlGn"))) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)  + 
  theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Smoothed Map of Black-to-White HIV-Related Mortality Rate Ratio,
United States, 2020 (GLM)", x = "Ratio ", y = "", fill = "") +
  theme(plot.title= element_text(hjust = 0.5))

ggsave(here("results/mult_fit_map.png"), units="in", width=10, height=4, dpi=300)
```

Next, let's look at the map for the ratio of predicted to observed (P/O) values. 
```{r}
ggplot(df_sf_scale, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = predicted_observed_mult)) +
  scale_fill_gradientn(colors = rev(brewer.pal(9, "RdYlGn"))) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + 
  theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Map of Predicted Over Observed Black-to-White HIV-Related Mortality Rate Ratio,
United States, 2020 (GLM)", x = "", y = "", fill = "Ratio") +
  theme(plot.title= element_text(hjust = 0.5))
ggsave(here("results/mult_pred_map.png"), units="in", width=7, height=4, dpi=300)

ggplot(df_sf_scale, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = predicted_observed_car)) +
  scale_fill_gradientn(colors = rev(brewer.pal(9, "RdYlGn"))) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) + 
  theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Map of Predicted Over Observed Black-to-White HIV-Related Mortality Rate Ratio,
United States, 2020 (CAR)", x = "", y = "", fill = "Ratio") +
  theme(plot.title= element_text(hjust = 0.5))
ggsave(here("results/car_pred_map.png"), units="in", width=7, height=4, dpi=300)
``` 
We can see that the model fitted pretty well. A higher P/O ratio (more red values) indicates that the model overestimated the observed values, while a lower P/O ratio (more green values) indicates that the model underestimates the observed values. We can see that the model underestimated rates primarily in the non-coast north-western states and north central states for persons of Hispanic, non-Hispanic White, and non-Hispanic multiple racial/ ethnic backgrounds.

>>>>

df_fit <- readRDS(here("data/processed_data/df_fit_alt.rds"))
df_raw <- readRDS(here("data/processed_data/race_alt.rds"))
```

I do a bit more cleaning to the data to tidy things up and create some new variables
```{r}
df_fit <- df_fit %>%
  mutate(region = tolower(State)) %>% 
  replace(is.na("Ratio Observed/ Expected"), 0) %>% 
  mutate(Exp_DR = `Expected Death Cases`/Population *1e5,
               OE_mortality_ratio = `Death Rate (estimated)`/Exp_DR,
         dykes_chi = ((`Death Cases`- `Expected Death Cases`)/ sqrt(`Expected Death Cases`)),

         ## GLM
         predicted_observed_mult = mult_fit/`MRR`,
         predicted_observed_mult = ifelse(!is.finite(predicted_observed_mult), 0, predicted_observed_mult),
         observed_predicted_mult = `Death Cases`/mult_fit,
         observed_predicted_mult = ifelse(!is.finite(observed_predicted_mult), 0, observed_predicted_mult),
         ## CAR
         predicted_observed_car = car_fit/`MRR`,
         predicted_observed_car = ifelse(!is.finite(predicted_observed_car), 0, predicted_observed_car),
         observed_predicted_car = `Death Cases`/car_fit,
         observed_predicted_car = ifelse(!is.finite(observed_predicted_car), 0, observed_predicted_car),
         ## POISSON
         predicted_observed_pois = pois_fit/`MRR`,
         predicted_observed_pois = ifelse(!is.finite(predicted_observed_pois), 0, predicted_observed_pois),
         observed_predicted_pois = `Death Cases`/pois_fit,
         observed_predicted_pois = ifelse(!is.finite(observed_predicted_pois), 0, observed_predicted_pois),
         ##N NEGATIVE BINOMIAL
         predicted_observed_nb = nb_fit/`MRR`,
         predicted_observed_nb = ifelse(!is.finite(predicted_observed_nb), 0, predicted_observed_nb),
         observed_predicted_nb = `Death Cases`/nb_fit,
         observed_predicted_nb = ifelse(!is.finite(observed_predicted_nb), 0, observed_predicted_nb))
```

Next, I will merge my dataframe contain the variables of interest with the spatial object. 
```{r}
df_sf_fit<- merge(states,df_fit, by = "region")
# This next step is import to ensure that no "holes" appear in the map.
df_sf_fit <- df_sf_fit[order(df_sf_fit$order), ]
```

Now I'm ready to start mapping! I'll begin by looking at the map of the raw HIV death rate across to contiguous US by selected race/ ethnicity.
```{r}
ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Death Rate (estimated)`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race)  + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "HIV Mortality Rate (raw) by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Death rate per 100,000") +
  scale_fill_distiller(palette = "RdYlGn") +
  theme(plot.title= element_text(hjust = 0.5))

@@

```{r}
ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`MRR`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio in the United States, 2020", x = "", y = "", fill = "Ratio") +
  scale_fill_distiller(palette = "RdYlGn") +
  theme(plot.title= element_text(hjust = 0.5))

@@

```{r}
map_1 <- ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Percent Living in Poverty`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Percent Living in Poverty, 2020", x = "", y = "", fill = "Percent")  +
  theme(plot.title= element_text(hjust = 0.5))

```

```{r}
map_2 <-ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Median Household Income`/1e4)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Median Household Income, 2020", x = "", y = "", fill = "Per $10,000")  +
  theme(plot.title= element_text(hjust = 0.5))

```

```{r}
map_3 <- ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Percent Uninsured Under 65 Years Old`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Percent Uninsured Under 65 Years Old, 2020", x = "", y = "", fill = "Percent") +
  theme(plot.title= element_text(hjust = 0.5))

```

```{r}
map_4 <- ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Percent Without an Internet subscription`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Percent Without an Internet Subscription, 2020", x = "", y = "", fill = "Percent") +
  theme(plot.title= element_text(hjust = 0.5))

```

```{r}
library(gridExtra)
grid_map<- grid.arrange(map_2, map_1, map_3, map_4, ncol = 2, nrow = 2)

ggsave(here("results/grid_map.png"), grid_map, units="in", width=11, height=10, dpi=300)
```

@@

Next, let's map a probability map based on where we'd expect deaths if risk of HIV-related death was consistent across the US.
```{r}
ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = Exp_DR)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race) +
  scale_fill_viridis_c() + theme_void() + 
  theme(strip.text = element_text(size = 14),
    strip.background = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.text = element_blank(),
        legend.position = "bottom") +
  labs(title = "NULL - Expected HIV Mortality Rate by Race/ Ethnicity
       in the United States, 2020", x = "", y = "", fill = "Death Rate (per 100,000)") +
  theme(plot.title= element_text(hjust = 0.5))
```

Next, let's using Dykes and Unwin's method of smooth map using a chi square value. 
```{r}
ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = dykes_chi)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race)  + theme_void() + 
  theme(strip.text = element_text(size = 14),
        strip.background = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.text = element_blank(),
        legend.position = "bottom") +
  labs(title = "Dykes and Unwin's Chi Square by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Chi Square") +
  scale_fill_distiller(palette = "RdYlGn") +
  theme(plot.title= element_text(hjust = 0.5))

@@

```{r}
##Smoothed map using fitted values from best model (CAR)
car_smooth <- ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = car_fit)) +
  scale_fill_gradientn(colors = rev(brewer.pal(9, "RdYlGn"))) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)  + 
  theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Smoothed Map of 
Black-to-White HIV-Related MRR (CAR)
United States, 2020", x = "Ratio", y = "", fill = "") +
  theme(plot.title= element_text(hjust = 0.5))

```

```{r}
##Smoothed map using fitted values from MULT (only marginally worse than CAR)
mult_smooth <-ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = mult_fit)) +
  scale_fill_gradientn(colors = rev(brewer.pal(9, "RdYlGn"))) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)  + 
  theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Smoothed Map of 
Black-to-White HIV-Related MRR (GLM)
United States, 2020", x = "Ratio ", y = "", fill = "") +
  theme(plot.title= element_text(hjust = 0.5))

grid_smooth <- grid.arrange(car_smooth, mult_smooth, ncol = 2, nrow = 1)
ggsave(here("results/grid_smooth.png"), grid_smooth, units="in", width=10, height=5, dpi=300)
```

Next, let's look at the map for the ratio of predicted to observed (P/O) values. 
```{r}
pred_mult <- ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = predicted_observed_mult)) +
  scale_fill_gradientn(colors = rev(brewer.pal(9, "RdYlGn"))) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + 
  theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Predicted Over Observed
Black-to-White HIV-Related MRR
United States, 2020 (GLM)", x = "", y = "", fill = "Ratio") +
  theme(plot.title= element_text(hjust = 0.5))

pred_car <-ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = predicted_observed_car)) +
  scale_fill_gradientn(colors = rev(brewer.pal(9, "RdYlGn"))) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) + 
  theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Predicted Over Observed
Black-to-White HIV-Related MRR
United States, 2020 (CAR)", x = "", y = "", fill = "Ratio") +
  theme(plot.title= element_text(hjust = 0.5))
```

```{r}
grid_pred <- grid.arrange(pred_car, pred_mult, ncol = 2, nrow = 1)

ggsave(here("results/grid_pred.png"), grid_pred, units="in", width=10, height=5, dpi=300)
```
We can see that the model fitted pretty well. A higher P/O ratio (more red values) indicates that the model overestimated the observed values, while a lower P/O ratio (more green values) indicates that the model underestimates the observed values. We can see that the model underestimated rates primarily in the non-coast north-western states and north central states for persons of Hispanic, non-Hispanic White, and non-Hispanic multiple racial/ ethnic backgrounds.

### 2023-04-30 products/manuscript/manuscript.qmd
Auto stash before revert of "fixed pca and models"

In order to assess spatial associations, this analysis was limited to the contiguous United States. Spatial object of the contiguous United States was obtained from library(maps) in R and a list of neighboring contiguous United States was obtained from a public repository on [GitHub](https://github.com/ubikuity/List-of-neighboring-states-for-each-US-state). [@ubikuity2022]

@@

Results of the PCA shows that the first principal component contributed for 91.7 % of variation between predictors. The variables `Percent Uninsured Under 65 Years Old` explained the least amountof variation. Ultimately, only `Median Household Income`, `Percent Living in Poverty`, `Percent Living with Food Insecurity`, and `Percent Without an Internet Subscription` were included as predictors in the full model.

@@

## Software
```{r}
sessionInfo() 
```
{{< pagebreak >}}

>>>>

In order to assess spatial associations, this analysis was limited to the contiguous United States. Spatial object of the contiguous United States was obtained from `library(maps)` in R and a list of neighboring contiguous United States was obtained from a public repository on [GitHub](https://github.com/ubikuity/List-of-neighboring-states-for-each-US-state). [@ubikuity2022]

@@

Results of the PCA shows that the first principal component contributed for 91.7 % of variation between predictors. All other components were excluded as they explained less than 20% of variation. The variables `Percent Uninsured Under 65 Years Old` explained the least amount of variation. Ultimately, only `Median Household Income`, `Percent Living in Poverty`, `Percent Living with Food Insecurity`, and `Percent Without an Internet Subscription` were included as predictors in the full model.

@@

## Software

```{r}
sessionInfo() 
```
